---
title: "MAT 5101 - Problem Set 2"
subtitle: "(Group 1)"  
author:
- Wayne Matthew Dayata
- Ivan Ric Woogue
- Raymond Aya-ay
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r include=FALSE}
require('SciViews')
require('PolynomF')
require('Deriv')
library('matlib')

```


```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
  require(PolynomF)
```
## Problem 1

Use the bisection method with a hand calculator or computer to find the real root 
of $x^3 - x - 3 = 0$. Use an error tolerance of $\epsilon = 0.0001$. 
Graph the function $f(x) = x^3 - x - 3$ and label the root.

```{r}
func = function(x){ x^3 - x - 3 }
```
$f(1) = -3$, $f(2) = 3$. As defined by the intermediate value theorem, $f(1)f(2) < 0$, and $f(x)$ is continuous. So there is a root within the interval $[1,2]$.

Solving using bisection method:

```{r echo=TRUE}
df = data.frame(matrix(ncol=7))
colnames(df) = c("#","a","b","c","f(b)","f(c)","f(b)*f(c)"); count = 0

bisect = function(f,a,b,epsilon){
  while(TRUE){
    count <<- count + 1
    c = (a+b)/2
    df[nrow(df)+1,] <<- list(count,a,b,c,f(b),f(c),f(b)*f(c)) # output
    if (abs(b-c)<=epsilon) {df <<- df[-1,]; return(c)}
    if (f(b)*f(c)<=0) {a=c} else {b=c}
  }
}

root = bisect(func,1,2,0.0001)
print(df, row.names = FALSE)
root # bisection result
uniroot(func,c(1,2))$root #uniroot, using Newton's Method
```

```{r echo=TRUE}
curve(func, xlim=c(-5,5), ylim=c(-10,10), col="red", lwd=3)
abline(h = 0, lty = 2)
abline(v = 0, lty = 2)
points(root, 0, pch=19, col="green")
text(root, 0, "x = 1.671692", pos = 3)
```

## Problem 2

The function $f(x) = -3x^3 + 2e^{x^2/2} - 1$ has values of zero near $x = -0.5$ 
and $x = 0.5$.

a. What is the derivative of $f(x)$?

$$f'(x) = 2x\mathrm{e}^\frac{x^2}{2}-9x^2$$
$$f'(x) = x\cdot\left(2\mathrm{e}^\frac{x^2}{2}-9x\right)$$
```{r echo=TRUE}
func = function(x){ (-3*x^3) + 2*exp(x^2/2)-1 }
dpdx = Deriv(func,"x")
dpdx
```


b. If you begin Newton’s method at $x = 0$, which root is reached? How many iterations to achieve an error less than $10^{-5}$?

Note: Since $f'(0) = 0$, using 0 as the starting point will yield an error when substituted in the formula for getting the estimated roots using Newton's method. We'll change our starting point to 0.5 instead.

```{r echo=TRUE}

library(Deriv)

func = function(x){ (-3*x^3) + 2*exp(x^2/2)-1 }
dpdx = Deriv(func,"x")

df = data.frame(matrix(ncol=6))
colnames(df) = c("#", "x_n", "x_n+1", "f(x_n)", "df(x_n)", "error")
count = 0
newtons = function(f, d_f, init_estimate, tolerance) {
  xn = init_estimate
  while(TRUE) {
    count <<- count + 1
    xn1 = xn - (f(xn) / d_f(xn))

    error = abs(xn1 - xn)
    
    df[nrow(df)+1, ] <<- list(count, xn, xn1, f(xn), d_f(xn), error)

    if(error <= tolerance) {
      df <<- df[-1,]
      return(xn1)
    }

    xn = xn1
  }
}

root = newtons(func, dpdx, 0.5, 0.00001)

print(df, row.names = FALSE)
print(paste("Root: ", root))
print(paste("Number of iterations:",count))

#Verifying solution with uniroot()
paste("Using uniroot(): ", uniroot(func,c(0,1))$root)

```

c. Begin Newton’s method at another starting point to get the other zero.

Based on the graph from #3, the other root of $f(x)$ is located to the right of the first obtained root.
With this, we can test with $x = 3.$

```{r echo=TRUE}

func = function(x){ (-3*x^3) + 2*exp(x^2/2)-1 }
dpdx = Deriv(func,"x")

df = data.frame(matrix(ncol=6))
colnames(df) = c("#", "x_n+1", "x_n", "f(x_n)", "df(x_n)", "error")
count = 0
newtons = function(f, d_f, init_estimate, tolerance) {
  xn = init_estimate
  while(TRUE) {
    count <<- count + 1
    xn1 = xn - (f(xn) / d_f(xn))

    error = abs(xn1 - xn)
    
    df[nrow(df)+1, ] <<- list(count, xn1, xn, f(xn), d_f(xn), error)

    if(error <= tolerance) {
      return(xn1)
    }

    xn = xn1
  }
}

root = newtons(func, dpdx, 3, 0.00001)
root

#Verifying solution with uniroot()
uniroot(func,c(2.5,3.5))$root
```

## Problem 3

Use the function from no.$2$ and find the root using the secant method where $x_0 = 0$ and $x_1 = 1$. Use an error
tolerance of $\epsilon = 0.001$.

```{r echo=TRUE}

func = function(x){
  -3*x^3 + 2*exp(x^2/2) - 1
}


df2 = data.frame(matrix(ncol=5))
colnames(df2) = c("#","x_n-1","x_n","x_n+1","error")
count = 0

xa = uniroot(func,c(0,1))$root
xa

secant = function(f,x0,x1,tol){
  xn = x1
  xn0 = x0
  xn1 = 0
  while(abs(xa-xn1)>tol){
    count <<- count + 1
    xn1 = xn - f(xn) * (xn - xn0)/(f(xn)-f(xn0))
    df2[nrow(df2)+1,] <<- list(count,xn0,xn,xn1,round(abs(xn1-xa),5)) # output
    if(xn < xn1){
      xn0 = xn
      xn = xn1
    }else{
      xn0 = xn1
      xn = xn
    }
  }
  df2 <<- df2[-1,]
  return(xn1)
}

root = secant(func,0,1,0.001)
root
print(df2, row.names = FALSE)

```

```{r echo=TRUE}
curve(func, xlim=c(-5,5), ylim=c(-10,10), col="red", lwd=3)
abline(h = 0, lty = 2)
abline(v = 0, lty = 2)
points(root, 0, pch=19, col="green")
text(root, 0, "x = 0.8565782", pos = 3)
```

## Problem 4

Consider the system

$$\begin{aligned}
10.2x + 2.4y - 4.5z &= 14.067 \\
-2.3x - 7.7y + 11.1z &= -0.996 \\
-5.5x - 3.2y + 0.9z &= -12.645
\end{aligned}$$

a. Present the augmented matrix of the system.

The augmented matrix for the given system of equations is:

$$\left[ 
    \begin{array}{ccc|c}
        10.2 & 2.4 & -4.5 & 14.067 \\
        -2.3 & -7.7 & 11.1 & -0.996 \\
        -5.5 & -3.2 & 0.9 & -12.645
    \end{array}
\right]$$

b. Solve the system using $Ax = LUx = Ly = b$ and round the
final answer to 4 decimal digits.

Let 
$$A = 
\begin{bmatrix}
        10.2 & 2.4 & -4.5 \\
        -2.3 & -7.7 & 11.1 \\
        -5.5 & -3.2 & 0.9
\end{bmatrix},
b = \begin{bmatrix}
        14.067 \\ -0.996 \\ -12.645
\end{bmatrix}.$$

We first factor A to LU. 

As shown below, we have LU = A which is equivalent to.

$$\begin{bmatrix}
1 & 0 & 0 \\
l_{21} & 1 & 0 \\
l_{31} & l_{32} & 1
\end{bmatrix}
\begin{bmatrix}
u_{11} & u_{12} & u_{13} \\
0 & u_{22} & u_{23} \\
0 & 0 & u_{33}
\end{bmatrix} = 
\begin{bmatrix}
        10.2 & 2.4 & -4.5 \\
        -2.3 & -7.7 & 11.1 \\
        -5.5 & -3.2 & 0.9
\end{bmatrix}$$

Multiplying the L and U matrices results to

$$\begin{bmatrix}
u_{11} & u_{12} & u_{13} \\
l_{21}u_{11} & l_{21}u_{12} + u_{22} & l_{21}u_{13} + u_{23} \\
l_{31}u_{11} & l_{31}u_{12} + l_{32}u_{22} & l_{31}u_{13} + l_{32}u_{23} + u_{33}
\end{bmatrix} = 
\begin{bmatrix}
        10.2 & 2.4 & -4.5 \\
        -2.3 & -7.7 & 11.1 \\
        -5.5 & -3.2 & 0.9
\end{bmatrix}$$

By equating the corresponding elements, we get:

$$\begin{aligned}
&u_{11} = a_{11} = 10.2 \\
&u_{12} = a_{12} = 2.4 \\
&u_{13} = a_{13} = -4.5 \\
&l_{21} = a_{21}/u_{11} = -2.3/10.2 = -0.225490 \\
&u_{22} = a_{22}-l_{21}u_{12} = -7.7-(-0.225490)(2.4) = -7.158824 \\
&u_{23} = a_{23}-l_{21}u_{13} = 11.1-(-0.225490)(-4.5) = 10.085295 \\
&l_{31} = a_{31}/u_{11} = -5.5/10.2 = -0.539216 \\
&l_{32} = (a_{32}-l_{31}u_{12})/u_{22} = (-3.2-(-0.539216)(2.4))/-7.158824 = 0.266228 \\
&u_{33} = a_{33}-l_{31}u_{13}-l_{32}u_{23} = 0.9-(-0.539216)(-4.5)-(0.266228)(10.085295) = -4.211460
\end{aligned}$$

Plugging the above values give the matrices:

$$L = 
\begin{bmatrix}
        1 & 0 & 0 \\
        -0.225490 & 1 & 0 \\
        -0.539216 & 0.266228 & 1
\end{bmatrix},
U = 
\begin{bmatrix}
        10.2 & 2.4 & -4.5 \\
        0 & -7.158824 & 10.085295 \\
        0 & 0 & -4.211460
\end{bmatrix}$$

In LUx = B, we let Ux = y to get Ly = b. We first solve for y in Ly = b and then solve for x in Ux = y where x represents the solutions $(x,y,z)^T$.

$$\begin{bmatrix}
        1 & 0 & 0 \\
        -0.225490 & 1 & 0 \\
        -0.539216 & 0.266228 & 1
\end{bmatrix}
\begin{bmatrix}
        y_1 \\ y_2 \\ y_3
\end{bmatrix} =\begin{bmatrix}
        14.067 \\ -0.996 \\ -12.645
\end{bmatrix}$$

The elements of y are:

$$\begin{aligned}
y_1 &= 14.067 \\
y_2 &= b_2-l_{21}y_1 = -0.996-(-0.225490)(14.067) = 2.175968 \\
y_3 &= b_3-l_{31}y_1-l_{32}y_2 = -12.645-(-0.539216)(14.067)-(0.266228)(2.175968) = -5.639152
\end{aligned}$$

Solving for x in Ux=y, we have:

$$\begin{bmatrix}
        10.2 & 2.4 & -4.5 \\
        0 & -7.158824 & 10.085295 \\
        0 & 0 & -4.211260
\end{bmatrix}
\begin{bmatrix}
        x_1 \\ x_2 \\ x_3
\end{bmatrix} =\begin{bmatrix}
        14.067 \\ 2.175968 \\ -5.639152
\end{bmatrix}$$

The elements of x are:

$$\begin{aligned}
z &= x_3 = y_3/u_{33} = -5.639152/-4.211260 = 1.339065 \\
y &= x_2 = (y_2-x_3u_{23})/u_{22} = (2.175968-(1.339065)(10.085295))/-7.158824 = 1.582508 \\
x &= x_1 = (y_1-x_3u_{13}-x_2u_{12})/u_{11} = (14.067-(1.339065)(-4.5)-(1.582508)(2.4))/10.2 = 1.597526
\end{aligned}$$

Therefore, $x = 1.597526, y = 1.582508, z = 1.339065$.

To verify, we multiply Ax to reobtain y.

$$Ax = \begin{bmatrix}
        10.2 & 2.4 & -4.5 \\
        -2.3 & -7.7 & 11.1 \\
        -5.5 & -3.2 & 0.9
\end{bmatrix}
\begin{bmatrix}
        1.597526 \\ 1.582508 \\ 1.339065
\end{bmatrix}
= \begin{bmatrix}
  14.066992 \\ -0.996000 \\ -12.645260
\end{bmatrix}
\approx \begin{bmatrix}
    14.067 \\ -0.996 \\ -12.645
\end{bmatrix} = b$$

```{r echo=TRUE}
A=matrix(c(10.2, 2.4, -4.5, -2.3, -7.7, 11.1, -5.5, -3.2, 0.9), nrow=3, byrow=T)
x_computed=matrix(c(1.597526, 1.582508, 1.339065), nrow=3, byrow=T)
A%*%x_computed # Multiplication
```
```{r echo=TRUE}
# LU Factorization Program (3x4 Augmented Matrix)

SolveLU = function(A,b){
  u11 = A[1,1]; u12 = A[1,2]; u13 = A[1,3]
  l21 = A[2,1]/u11
  u22 = A[2,2]-l21*u12
  u23 = A[2,3]-l21*u13
  l31 = A[3,1]/u11
  l32 = (A[3,2]-l31*u12)/u22
  u33 = A[3,3]-l31*u13-l32*u23

  L = matrix(c(1,0,0,l21,1,0,l31,l32,1),nrow=3,byrow=T)
  U = matrix(c(u11,u12,u13,0,u22,u23,0,0,u33),nrow=3,byrow=T)

  print("L:")
  print(L)
  
  print("U:")
  print(U)
  
  y = solve(L,b)
  print("y:")
  print(y)

  x = solve(U,y)
  print("x:")
  print(x)
}

A = matrix(c(10.2, 2.4, -4.5, -2.3, -7.7, 11.1, -5.5, -3.2, 0.9), nrow=3, byrow=T)
b = matrix(c(14.067, -0.996, -12.645), nrow=3, byrow=T)
SolveLU(A,b)
```

c. Find the residual vector based on the actual solution in solve().
```{r echo=TRUE}
x_true = solve(A,b)
x_true
```

Using the formula for residual vector $r = b-A\bar{x}$ (where $\bar{x}$ is the computed value of x), we have:

$$r = \begin{bmatrix}
        14.067 \\ -0.996 \\ -12.645
\end{bmatrix} - 
\begin{bmatrix}
        10.2 & 2.4 & -4.5 \\
        -2.3 & -7.7 & 11.1 \\
        -5.5 & -3.2 & 0.9
\end{bmatrix}
\begin{bmatrix}
        1.597526 \\ 1.582508 \\ 1.339065
\end{bmatrix} = 
\begin{bmatrix}
        0.0000081 \\ -0.0000001 \\ 0.0002601
\end{bmatrix}$$

Or through the alternate formula $r = A(x-\bar{x})$, we have:

$$r = \begin{bmatrix}
        10.2 & 2.4 & -4.5 \\
        -2.3 & -7.7 & 11.1 \\
        -5.5 & -3.2 & 0.9
\end{bmatrix}
\left(
\begin{bmatrix}
        1.597520 \\ 1.582419 \\ 1.339002
\end{bmatrix} - 
\begin{bmatrix}
        1.597526 \\ 1.582508 \\ 1.339065
\end{bmatrix}
\right) = 
\begin{bmatrix}
        0.0000081 \\ -0.0000001 \\ 0.0002601
\end{bmatrix}$$

In R, both solutions yield exactly the same residual vector.

```{r echo=TRUE}
r = b-A%*%x_computed
r 
r = A%*%(x_true-x_computed)
r
```

## Problem 5

Compute the Frobenius norm, maximum column sum, and maximum row sum of the matrix:

$$\begin{bmatrix}
10.2 & 2.4 & 4.5\\ 
-2.3 & 7.7 & 11.1\\ 
-5.5 & -3.2 & 0.9 
\end{bmatrix}$$

```{r}
A=matrix(c(10.2,2.4,4.5,-2.3,7.7,11.1,-5.5,-3.2,0.9), nrow = 3, byrow = T)
A
```

Forbenius norm: $$||A||_f=\sqrt{\sum_{i=1}^{m}\sum_{j=1}^{n}a_{ij}^2}$$ $$||A||_f=\sqrt{10.2^2+2.4^2+4.5^2+2.3^2+7.7^2+11.1^2+5.5^2+3.2^2+0.9^2}$$ $$||A||_f=\sqrt{359.14}$$ $$||A||_f=18.9509894201$$

```{r}
sqrt(sum(abs(A)^2))
```

Maximum column sum: $$||A||_1=max_{ 1\leq j\leq n}\sum_{i=1}^n|a_{ij}|$$ $$||A||_1=max((10.2+2.3+5.5),(2.4+7.7+3.2),(7.5+11.1+0.9))$$ $$||A||_1=max(18,13.3,16.5)$$ $$||A||_1=18$$

```{r}
max(colSums(abs(A)))
```

Maximum row sum:$$||A||_\infty=max_{ 1\leq j\leq n}\sum_{j=1}^n|a_{ij}|$$ $$||A||_\infty=max((10.2+2.4+4.5),(2.3+7.7+11.1),(5.5+3.2+0.9))$$ $$||A||_\infty=max(17.1,21.1,9.6)$$ $$||A||_\infty=21.1$$

```{r}
max(rowSums(abs(A)))
```

## Problem 6

Solve the system of equations given in no. 4, starting with the initial vector of $\left[0, 0, 0\right]$:

Since the given system of equations in number 4 is not diagonally dominant, it does not suffice the condition for the convergence of the approximations obtained by Jacobi or the Gauss-Seidel Method. To be diagonally dominant, it must satisfy the following conditions:

$$|A[1][1]| \geq |A[1][2]| + |A[1]A[3]|$$
$$|A[2][2]| \geq |A[2][1]| + |A[2]A[3]|$$
$$|A[3][3]| \geq |A[3][1]| + |A[3]A[2]|$$

The given matrix in number 4 is not diagonally dominant because the absolute of the 2nd and 3rd diagonally element is not greater than or equal to the sum of all the absolute elements in the corresponding row.

$$\begin{bmatrix}
10.2 & 2.4 & -4.5\\ 
-2.3 & -7.7 & 11.1\\ 
-5.5 & -3.2 & 0.9 
\end{bmatrix}$$

$$|-7.7| < |-2.3| + |11.1| \text{ or } 13.4$$
$$|0.9| < |-5.5| + |-3.2| \text{ or } 8.7$$


Therefore, we will be using a different matrix where it is diagonally dominant.

$$\left[\begin{array}{ccc|c}
10.7 & 3.9 & -2.2 & 14.067\\ 
-4.6 & 13.9 & -6.1 & -0.996\\ 
2.2 & -4.1 & -8.5 & -12.645
\end{array}
\right]$$

a. Solve using the Jacobi method with 2-digit precision.

```{r echo=TRUE}
library(matlib)

df = data.frame(matrix(ncol=4))
colnames(df) = c("#","x11","x21","x31") 
count = 0

A = matrix(c(10.7, 3.9, -2.2, 
             -4.6, 13.9, -6.1, 
             2.2, -4.1, -8.5), nrow=3, byrow=T)
A
L=matrix(c(0,0,0,-4.6,0,0,2.2,-4.1,0), nrow = 3, byrow = T)
L
D=matrix(c(10.7,0,0,0,13.9,0,0,0,-8.5), nrow = 3, byrow = T)
D
U=matrix(c(0,3.9,-2.2,0,0,-6.1,0,0,0), nrow = 3, byrow = T)
U
b = matrix(c(14.067, -0.996, -12.645), nrow=3, byrow=T)

print("Checking if L, D, U matches those in A")
A == L + D + U

SolveJacobi = function(L, D, U, b, x0, error) {
  count = 0
  repeat {
    x1=(-inv(D))%*%(L+U)%*%x0+inv(D)%*%b
    count <- count + 1
    df[nrow(df)+1,] <<- list(count, x1[1,1], x1[2,1], x1[3,1])
    if(
      # count == 8 ||
       (abs(x0[1,1] - x1[1,1]) <= error &&
       abs(x0[2,1] - x1[2,1]) <= error &&
       abs(x0[3,1] - x1[3,1]) <= error)) {
      df <<- df[-1,]
      return(x1)
      break;
    }
    x0 = x1
  }
}

PrintSideBySide = function(expected, actual) {
  df = data.frame(matrix(ncol=2))
  colnames(df) = c("Expected","Actual") 
  
  df[1,1] = expected[1,1]
  df[2,1] = expected[2,1]
  df[3,1] = expected[3,1]
  
  df[1,2] = actual[1,1]
  df[2,2] = actual[2,1]
  df[3,2] = actual[3,1]
  
  return(df)
}

x0 = matrix(c(0,0,0), nrow=3, byrow=T)
expected_ans = solve(A, b)
actual = SolveJacobi(L, D, U, b, x0, 0.001)

```


Using the formula $x^{n+1} = -D^{-1}(L+U)x^n+D^{-1}b$
```{r}
print(df, row.names = FALSE)
print(PrintSideBySide(expected_ans, actual), row.names=FALSE)
```


b. Solve using Gauss-Seidel method with 2-digit precision.

```{r echo=TRUE}
library(matlib)

df = data.frame(matrix(ncol=4))
colnames(df) = c("#","x11","x21","x31") 
count = 0

A = matrix(c(10.7, 3.9, -2.2, 
             -4.6, 13.9, -6.1, 
             2.2, -4.1, -8.5), nrow=3, byrow=T)
A

L=matrix(c(0,0,0,-4.6,0,0,2.2,-4.1,0), nrow = 3, byrow = T)
L
D=matrix(c(10.7,0,0,0,13.9,0,0,0,-8.5), nrow = 3, byrow = T)
D
U=matrix(c(0,3.9,-2.2,0,0,-6.1,0,0,0), nrow = 3, byrow = T)
U
b = matrix(c(14.067, -0.996, -12.645), nrow=3, byrow=T)

print("Checking if L, D, U matches those in A")
A == L + D + U

SolveGaussSeidel = function(L, D, U, b, x0, error) {
  count = 0
  repeat {
    count <- count + 1
    x1=-inv(L+D)%*%U%*%x0+inv(L+D)%*%b
    df[nrow(df)+1,] <<- list(count, x1[1,1], x1[2,1], x1[3,1])
    if(
       (abs(x0[1,1] - x1[1,1]) <= error &&
       abs(x0[2,1] - x1[2,1]) <= error &&
       abs(x0[3,1] - x1[3,1]) <= error)) {
      df <<- df[-1,]
      return(x1)
      break;
    }
    x0 = x1
  }
}
 
x0 = matrix(c(0,0,0), nrow=3, byrow=T)
expected_ans = solve(A, b)
actual = SolveGaussSeidel(L, D, U, b, x0, 0.001)

```

Using the formula $x^{n+1} = -(L+D)^{-1}Ux^n+(L+D)^{-1}b$
```{r }
print(df, row.names = FALSE)
print(PrintSideBySide(expected_ans, actual), row.names=FALSE)

```


c. Solve for $\bar{e}$ if the computed solution is $\bar{x} = (1.2515925, 0.9386049, 1.3588499)^T$.
The true solution is $x = (1.251825, 0.938910, 1.358763)^T$

```{r echo=TRUE}

library(matlib)

A = matrix(c(10.7, 3.9, -2.2, 
             -4.6, 13.9, -6.1, 
             2.2, -4.1, -8.5), nrow=3, byrow=T)
A
b = matrix(c(14.067, -0.996, -12.645), nrow=3, byrow=T)
b
x_bar = matrix(c(1.2515925, 0.9386049, 1.3588499))
x_bar

r = b-A%*%x_bar

e_bar = inv(A)%*%r

x_barcorrected=x_bar+e_bar

```


$r = b-A\bar{x}$
```{r}
r
```


$\bar{e} = A^{-1}r$
```{r}
e_bar
```


$x = \bar{x} + \bar{e}$
```{r}
x_barcorrected
```


The sum of the computed solution $\bar{x}$ and the error $\bar{e}$ is equal to the true solution. Therefore, the error is correct. The $\bar{e}$ is found below.

```{r}
e_bar
```













