---
title: "Trapezoidal, Simpson, and Romberg Integration Rules"
author: "Hannah B. Labana, Jomar M. Leaño, Christian Anthony C. Stewart"
date: "2024-02-29"
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes: |
  \usepackage{graphics}
  \graphicspath{ {figures/} }
  \usepackage{array}
  \usepackage{amsmath}
  \usepackage{xcolor}
  \usepackage{bigints}
---

```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
  require(PolynomF)
  require('Deriv')
  library('pracma')
```

$\int_1^2 sin(x^2) d x$

## 1. Trapezoid Rule

```{r}
f = function(x) {sin(x^2)}

# Define the limits of integration
a = 1
b = 2

# Number of intervals
n = 18

# Width of each interval
h = (b - a) / n

# Trapezoid rule formula
int = (h / 2) * (f(a) + 2 * sum(sapply(1:(n - 1), function(i) f(a + i * h))) + f(b))

actual = integrate(f, lower = 1, upper = 2)
```

```{r echo=FALSE}
paste("Estimated value: ", int)
paste("Actual value: ", actual[1])
paste("Error: ", abs(actual$value-int))
```
for $n = 18$, we are guaranteed

$$\left\lvert\int_{1}^{2}sin(x^2) - M_n\right\rvert \le 0.001$$

take note that this took around 18 intervals between 1 and 2 which is a lot.

## 2. Simpson's Rule

```{r}
f = function(x) {sin(x^2)}

# Define the limits of integration
a = 1
b = 2

# Number of intervals
n = 6

simpsons_rule <- function(f, a, b, n) {
  if (n %% 2 != 0) {
    stop("Number of subintervals must be even.")
  }
  
  h <- (b - a) / n
  integral <- f(a) + f(b)  # endpoints
  
  # Odd indexed points
  odd_sum <- sum(sapply(seq(1, n, by = 2), function(i) f(a + i * h)))
  # Even indexed points
  even_sum <- sum(sapply(seq(2, n-1, by = 2), function(i) f(a + i * h)))
  
  integral <- integral + 4 * odd_sum + 2 * even_sum
  integral <- integral * h / 3
  
  return(integral)
}

actual = integrate(f, lower = 1, upper = 2)
```

```{r echo=FALSE}
paste("Estimated value: ", int <- simpsons_rule(f, a, b, n))
paste("Actual value: ", actual[1])
paste("Error: ", abs(actual$value-int))
```

for $n = 6$, we are guaranteed

$$\left\lvert\int_{1}^{2}sin(x^2) - M_n\right\rvert \le 0.001$$

take note that this took around 6 intervals between 1 and 2 which is significantly more efficient compared to Trapezoid Rule which took 18 intervals.

### Romberg Integration
```{r}
romberg_integration = function(num_of_iter, func, lower_bound, upper_bound) {
    deltax_array = array(data = NA, dim = num_of_iter+1)
    T_array = array(data = NA, dim = num_of_iter+1)
    Tx_array = array(data = NA, dim = num_of_iter+1)
    
    ptr = 1
    prev = 1
    after = 2
    deltax = (upper_bound - lower_bound) / 2
    
    while(ptr < num_of_iter+2) {
        deltax_array[ptr] = deltax
        num_of_subintervals = 2^num_of_iter
        temp = 0
        x = 1
        
        while(x < 2^ptr) {
            temp = temp + 2 * func(lower_bound + x * deltax)
            x = x + 1
        }
        
        T_d = (deltax/2) * (func(lower_bound) + temp + func(upper_bound))
        T_array[ptr] = T_d
        
        if(ptr > 1) {
            Tx = ((2^(2*prev)) * T_array[after] - T_array[prev]) / (2^(2*prev) - 1)
            Tx_array[prev] = Tx
            prev = prev + 1
            after = after + 1
        }
        
        ptr = ptr + 1
        deltax = deltax / 2
    }
    
    result_df <- data.frame(deltax = deltax_array, T = T_array, Tx = Tx_array + 1)
    result_df$deltax <- deltax_array
    result_df$T <- T_array
    result_df$Tx <- Tx_array
    
    return(list(Tx_array, result_df))
}

func_to_integrate = function(x) {
    return(sin(x^2))
}

result = romberg_integration(6, func_to_integrate, 1, 2)
answer = result[1]
result_dataframe = result[2]
```

```{r echo="FALSE"}
result_dataframe
```

```{r echo="FALSE"}
answer
```
We can see the estimated answer at the 6th iteration = 0.4944894
$\\$If we use the pracma library, it gives 0.4945082 with 6 iterations.
```{r}
f <- function(x) {
  sin(x^2)
}

rom <- romberg(f, 1, 2)
rom
```
Error = |Expected - Actual|
$\\$
```{r}
0.4945082 - 0.4944894
```

```{r}
integrate(f,1,2)
```