---
title: "Problem Set 1"
author: "Hannah Labana, Jomar Leaño, Christian Stewart"
output: pdf_document
date: "2024-01-31"
header-includes: |
  \usepackage{graphicx}
  \graphicspath{ {figures/} }
  \usepackage{array}
  \usepackage{amsmath}
  \usepackage{xcolor}
  \usepackage{bigints}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require('SciViews')
require('PolynomF')
require('Deriv')
```

## Problem Set 1

1.Prove that $lim_{x\rightarrow -1}{2x}+1=-1.$

Note: For any given $\epsilon > 0$, there exists a $\delta > 0$ such that $|f (x) - f (x_0)| < \epsilon$ whenever $|x - x_0| < \delta$.

  Draft:

$$|2x+1 - (-1)| < \epsilon$$
$$|2x+2| < \epsilon$$
$$|2||x-(-1)|<\epsilon$$
$$|x-(-1)|< \frac{\epsilon}{|2|} $$
  **End of Draft:** $\delta\leq\frac{\epsilon}{2}$

  **Start of Formal Solution:**
$\newline$ Given we have $\epsilon$ > 0. $\newline$ Let $\delta \leq \frac{\epsilon}{2}$ shows that when $|x-(-1)|< \delta$, then $|2x+1-(-1)|<\epsilon$

  $$|x-(-1)| < \delta$$
  $$|x-(-1)| < \frac{\epsilon}{2}$$
  $$|2||x-(-1)|<\epsilon$$
  $$|2x+2| < \epsilon$$
  $$|2x+1-(-1)| < \epsilon$$
  
Therefore, $lim_{x\rightarrow -1}{2x}+1=-1$.

2.Determine all the numbers c which satisfy the conclusions of the Mean Value Theorem for the following function and graph using R with the point/s identified. $f(x)=x^3-4x^2-2x-5$ on $[-10,10]$

Let $f(x)$ be continuous for $a \leq x \leq b$, and let it be differentiable for $a < x < b$. Then there is at least one point $\xi$ in $(a, b)$ for which $$f(b) - f(a) = f'(\xi)(b - a)$$
$$f\prime(\xi) = \frac{f(b) - f(a)}{b - a}$$
```{r echo=FALSE}
x = polynom()
p = x^3 - 4*x^2 - 2*x - 5

# Intervals
a = -10
b = 10

# Defining formula for Mean Value Theorem
f <- function(x) {
return(p(x))
}

# Solving the Mean Value Theorem
c <- (f(b) - f(a)) / (b - a)

```

```{r echo=FALSE}
# Solving derivative
dpdx = deriv(p)
dpdxm <- dpdx - c
dpdxm

# Getting the roots
dpdxm_zeros = solve(dpdxm)
dpdxm_zeros
```

Since f(x) is a polynomial, it is both continuous and differentiable, so $f'(x) = 3x^2 - 8x - 2$.
we then replace x with the values provided which was $[-10,10]$


$$f(x) = x^3 - 4x^2 - 2x - 5$$
$$f(-10) = (-10)^3 - 4(-10)^2 - 2(-10) - 5$$
$$= -1000 - 400 + 20 - 5$$
$$= f(x) = -1385$$

$$f(x) = x^3 - 4x^2 - 2x - 5$$
$$f(10) = 10^3 - 4(10)^2 - 2(10) - 5$$
$$= 1000 - 400 - 20 - 5$$
$$= f(x) = 575$$
we then use this equation for the calculation of the point we need to find
$$f\prime(\xi) = \frac{f(b) - f(a)}{b - a}$$
$$a = -1385, b = 575$$
$$3\xi^2 - 8\xi - 2 = \frac{575-(-1385)}{10-(-10)}$$

$$3\xi^2 - 8\xi - 2 = 98$$

$$3\xi^2 - 8\xi - 100 = 0$$
we can now solve for the roots of the equation using the following

$$x = \frac{-b \pm \sqrt{b^2-4(a)(c)}}{2a}$$
$$x = \frac{-(-8) \pm \sqrt{(-8)^2-4(3)(-100)}}{2(3)}$$
$$x = \frac{8 \pm \sqrt{64-12(-100)}}{6}$$
$$x = \frac{8 \pm \sqrt{64+1200}}{6}$$
$$x = \frac{8 \pm \sqrt{1264}}{6}$$
$$x_0 \approx  7.258796, x_1 \approx  -4.592130$$
the solutions on $[-10, 10]$ are $x_0 \approx  7.258796, x_1 \approx  -4.592130$

```{r echo=FALSE}
# get the parallel lines - slope and given point
m = dpdx(dpdxm_zeros[1])
# get the y-intercept (b)
# y - y0 = m (x - x0) -> point slope form
# y = m (x - x0) + y0
# y = mx - mx0 + y0 -> y-intercept form (y=mx+b)
# b = -mx0 + y0 
x1 = dpdxm_zeros[1]
y1 = p(dpdxm_zeros[1])
x2 = dpdxm_zeros[2]
y2 = p(dpdxm_zeros[2])
b1 = -m*x1 + y1 # intercept of left parallel line
b2 = -m*x2 + y2 # intercept of right parallel line

curve(p, -10, 10, col="black", lwd=3)
curve(dpdxm, -10, 10, col="blue", lwd=3, add=T)

points(dpdxm_zeros[1], p(dpdxm_zeros[1]), pch=19, col="red")
points(dpdxm_zeros[2], p(dpdxm_zeros[2]), pch=19, col="red")

points(dpdxm_zeros[1], dpdxm(dpdxm_zeros[1]), pch=19, col="green")
points(dpdxm_zeros[2], dpdxm(dpdxm_zeros[2]), pch=19, col="green")
abline(h=0, lty=2)

# plotting parallel lines
abline(a=b1, b=m, pch=5, lty=2, col="gray")
abline(a=b2, b=m, pch=5, lty=2, col="gray")

# original slope
m_orig = c # (f(b)-f(a))/(b-a)
b3 = (f(b) + f(a))/2
abline(a=b3, b=m_orig, pch=15, lty=2, lwd=2, col="orange")

text(dpdxm_zeros[1], dpdx(dpdxm_zeros[1]), "f'(-4.5921)", pos = 3)
text(dpdxm_zeros[2], dpdx(dpdxm_zeros[2]), "f'(7.2587)", pos = 1)

```
3.Find the point c that satisfies the mean value theorem for integrals on the interval [-1,1]. The function is $f(x)=3e^2$

**Using the Mean Value Theorem**
           $$Mean Value Theorem = f'\left(c\right) = \frac{{f\left(b\right)-f\left(a\right)}}{{b - a}}$$
We then derive:
 $$f'(x) = \frac{d}{dx} (2e^x)$$
 $$      = 2e^x$$
 $$f(1)  = 2e^1 = 2e$$
 $$f(-1) = 2e^{-1}$$
Average rate of change: 
 $$\frac{f(1)-f(-1)}{1-(-1)}$$ 
 $$=\frac{2e^1-2e^{-1}}{2}$$ 
 $$= e - \frac{1}{e}$$
Finding c:
 $$f'(c) = e - \frac{1}{e}$$
 $$2e^c  = e - \frac{1}{e}$$
 $$e^c   = \frac{e-\frac{1}{e}}{2}$$
 $$c   = ln(\frac{e-\frac{1}{e}}{2})$$
 $$\therefore c   \approx ln(\frac{e-\frac{1}{e}}{2})$$
4.Consider the function $f(x) = cos(\frac{x}{2})$
  $\newline$
  a. Find the fourth Taylor polynomial for $f$ at $x=\pi$
  $$f(x) = \cos(\frac{x}{2}), f(\pi) = 0$$
  $$f^1(x) = -\frac{\sin(\frac{x}{2})}{2}, f^1 (\pi) = -\frac{1}{2}$$, which is -0.5
  $$f^2(x) = -\frac{\cos(\frac{x}{2})}{4}, f^2 (\pi) = 0$$
  $$f^3(x) = \frac{\sin(\frac{x}{2})}{8}, f^3 (\pi) = -\frac{1}{8}$$, which is 0.0125
  $$f^4(x) = \frac{\cos(\frac{x}{2})}{16}, f^4 (\pi) = 0$$
  $$f^5(x) = -\frac{\sin(\frac{x}{2})}{32}, f^5 (\pi) = -\frac{1}{32}$$, which is -0.03125
    to find the fourth Taylor Polynomial, we use the following equation:
  
  $$p_4 = f(x_0) + \frac{f^1(x_0)}{1!}(x-x_0) + \frac{f^2(x_0)}{2!}(x-x_0)^2 + \frac{f^3(x_0)}{3!}(x-x_0)^3 + \frac{f^4(x_0)}{4!}(x-x_0)^4$$
  $$p_4 = 0 + \frac{-\frac{1}{2}}{1!}(x-\pi) + \frac{0}{2!}(x-\pi)^2 + \frac{\frac{1}{8}}{3!}(x-\pi)^3 + \frac{0}{4!}(x-\pi)^4$$
  $$p_4 = \frac{-\frac{1}{2}}{1!}(x-\pi)+ \frac{\frac{1}{8}}{3!}(x-\pi)^3$$
  $\newline$b. Use the fourth Taylor polynomial to approximate $cos(\frac{\pi}{2})$.
   $\newline$To find the fourth Taylor Polynomial, we use the following equation:
  
  $$p_4 = f(x_0) + \frac{f^1(x_0)}{1!}(x-x_0) + \frac{f^2(x_0)}{2!}(x-x_0)^2 + \frac{f^3(x_0)}{3!}(x-x_0)^3 + \frac{f^4(x_0)}{4!}(x-x_0)^4$$
  $$p_4 = 0 + \frac{-\frac{1}{2}}{1!}(\frac{\pi}{2}-\pi) + \frac{0}{2!}(\frac{\pi}{2}-\pi)^2 + \frac{\frac{1}{8}}{3!}(\frac{\pi}{2}-\pi)^3 + \frac{0}{4!}(\frac{\pi}{2}-\pi)^4$$
  $$p_4 = -\frac{1}{2}(-\frac{\pi}{2}) + \frac{1}{48}(-\frac{\pi}{2})^3$$
  $$p_4 = \frac{\pi}{4} - \frac{\pi^3}{324}$$
  
  $$Actual Value = 0.7071068$$
  $$Estimated Value \approx 0.7046527$$
```{r echo=FALSE}
f = function(x){cos(x/2)}
p1 = function(x){(-1/2)*(x-pi)}
p4 = function(x){(-1/2)*(x-pi)+(1/48)*((x-pi)^3)}
curve(f, -10, 10, col="purple", lwd=3)
curve(p1, -10, 10, col="orange", lwd=2, add=T, lty=2)
curve(p4, -10, 10, col="red", lwd=2, add=T, lty=2)

text(-3.1, 0.8, "cos(x/2)", pos = 1, col="purple")
text(1.5, 0.8, "p1", pos = 4, col="orange")
text(0, 0.6, "p3", pos = 1, col="red")

points(pi, f(pi), pch=19, lwd=4, col="green")
points(pi/2, p4(pi/2), pch=19, lwd=2, col="red")
points(pi/2, f(pi/2), pch=19, lwd=2, col="orange")
```
  
  $\newline$c. Use the fourth Taylor polynomial to bound the error
   By Taylor's Theorem with remainder, there exists a $\varepsilon$ in the interval $(\frac{\pi}{2},\pi)$ such that the remainder when approximating $cos(\pi/4)$ by the fourth Taylor polynomial satisfies

  $$R_4 \left(\frac{\pi}{2}\right) = \frac{f^5(\frac{\pi}{2})}{5!} (x-\pi)^5$$  
  We do not know the exact value of $c$, we find an upper bound on $R_4 \left(\frac{\pi}{2}\right)$ by determining the maximum value of $|f^5(x)|$ on the interval $(\frac{\pi}{2},\pi)$.
  
  $$-0.02209709, -0.03125000$$
```{r echo=FALSE}
f=function(x){-sin(x/2)/32}
f(c(pi/2,pi))
curve(f,pi/2,pi)
```
  
  $$R_5 \left(\frac{\pi}{2}\right) \leq \frac{-0.03125000}{5!} (\frac{\pi}{2}-\pi)^5 \approx 0.002490395$$
  $$cos \left(\frac{\pi}{4}\right) \approx p_4\left(\frac{\pi}{2}\right) + R_4 \left(\frac{\pi}{2}\right)
= 0.7046527 + 0.002490395$$
  $$cos \left(\frac{\pi}{4}\right) \approx 0.7071431$$
```{r echo=FALSE}
0.7046527 + 0.002490395 
abs(cospi(1/4) - (0.7046527 + 0.002490395))
```
  
5.If fl(x) is the machine approximated number of a real number
x and $\epsilon$ is the corresponding relative error, then show that $fl(x) = (1 - \epsilon)x$

**Let:**
$\newline$	$fl(x) = Approximate Value$
$\newline$	$x     = Real Value$
$\newline$	$\epsilon     = Relative Error$

**Relative Error:**
$\newline$ 	$$\epsilon = |\frac{x - fl(x)}{x}|$$

**Then we solve for fl(x):**
	$$\epsilon|x|   = |x - fl(x)|$$
	$$fl(x)  = x - \epsilon|x|$$
	$$fl(x)  = (1 - \epsilon)x$$


6.For the following numbers $x$ and their corresponding approximations $x_A$, find the number of significant digits in $x_A$ with respect to $x$ and find the relative error.


$$Error = True Value - Approximate Value$$
$$Relative Error = Error/True Value$$

  a. $x=451.01, x_A=451.023$

the real number $x=451.01$ can be represented as $451.01=(-1)^0$ x $0.45101$ x $10^3$

we have $s=0, \beta=10^3, e=4, d_1=4, d_2=5, d_3=1, d_4=0, d_5=1$

Suppose that the true value $x=451.01$ and the approximate value $x_A=451.023$. Then the Error$(x_A)$ and the Relative Error $(x_A)$:

**Error = $451.01-451.023$ = $-0.013$**

**Relative Error = $\frac{451.01-451.023}{451.01}$ = $-0.00002882419$**

If $x_A$ is an approximation to $x$, then we say that $x_A$ approximates $x$ to $r$ significant $\beta$-digits if
$$\left|x-x_A\right| \leq \frac{1}{2} \beta^{s-r+1}$$
where $$\beta^s \leq |x|$$
Since $10^2<451.01=x$, $s=2$ and to determine r, we get
$$|x-x_A|=0.013 \leq \frac{1}{2}\times 10^{2-r+1}=\frac{1}{2}\times 10^{-1} = 0.05$$
$$2-r+1=-1$$
$$r=4$$
$$\left|451.01-451.023\right| \leq \frac{1}{2} 10^{2-4+1}$$
$$0.013 \leq \frac{1}{2} 0.1$$
$$0.013 \leq 0.05$$
which is true, so the approximate number has 4 significant digits and the relative error is $-0.00002882419$

  b. $x=-0.04518, x_A=-0.045113$

Suppose that the true value $x=-0.04518$ and the approximate value $x_A=-0.045113$. Then the Error$(x_A)$ and the Relative Error $(x_A)$:

**Error = $-0.04518-(-0.045113)$ = $0.000067$**

**Relative Error = $\frac{-0.04518-(-0.045113)}{-0.04518}$ = $0.001482957$**

If $x_A$ is an approximation to $x$, then we say that $x_A$ approximates $x$ to $r$ significant $\beta$-digits if
$$\left|x-x_A\right| \leq \frac{1}{2} \beta^{s-r+1}$$
where $$\beta^s \leq |x|$$
Since $10^{-2}<0.04518=x$, $s=-2$ and to determine r, we get
$$|x-x_A|=0.000067 \leq \frac{1}{2}\times 10^{-2-r+1}=\frac{1}{2}\times 10^{-3} = 0.0005$$
$$-2-r+1=-3$$ 
$$r=2$$ 
$$\left|-0.04518-(-0.045113)\right| \leq \frac{1}{2} 10^{-2-2+1}$$
$$0.000067 \leq \frac{1}{2} 10$$
$$0.000067 \leq 5$$
which is true, so the approximate number has 2 significant digits and the relative error is $0.001482957$

  c. $x=23.4604, x_A=23.4213$

Suppose that the true value $x=23.4604$ and the approximate value $x_A=23.4213$. Then the Error$(x_A)$ and the Relative Error $(x_A)$:

**Error = $23.4604-23.4213$ = $0.0391$**

**Relative Error = $\frac{23.4604-23.4213}{23.4604}$ = $0.001666638$**

If $x_A$ is an approximation to $x$, then we say that $x_A$ approximates $x$ to $r$ significant $\beta$-digits if
$$\left|x-x_A\right| \leq \frac{1}{2} \beta^{s-r+1}$$
where $$\beta^s \leq |x|$$
Since $10^1<23.4604=x$, $s=1$ and to determine r, we get
$$|x-x_A|=0.0391 \leq \frac{1}{2}\times 10^{1-r+1}=\frac{1}{2}\times 10^{-1} = 0.05$$
$$1-r+1=-1$$
$$r=2$$
$$\left|23.4604-23.4213\right| \leq \frac{1}{2} 10^{1-2+1}$$
$$0.0391 \leq \frac{1}{2} 1$$
$$0.0391 \leq 0.5$$
which is true, so the approximate number has 2 significant digits and the relative error is $0.001666638$

7.Find the condition number for the following functions
$\newline$ a.$f(x)=2x^2$
$\newline$The derivative of the equation is $f'(x) = 4x$
$$CN = \Bigg|\frac{f'(x)}{f(x)}x\Bigg|$$
$$=\Bigg|\frac{4x}{2x^2}x\Bigg|$$
$$=\frac{4}{2}$$
$$CN = 2$$

b.$f(x)=2\pi^x$
$\newline$The derivative of the equation is $f'(x) = 2\pi^x * ln\pi$
$$CN = \Bigg|\frac{f'(x)}{f(x)}x\Bigg|$$
$$ = \Bigg|\frac{2\pi^xln\pi}{2\pi^x}x\Bigg|$$
$$CN = |x|ln\pi$$

c.$f(x)=2b^x$
$\newline$The derivative of the equation is $f'(x) = 2b^xlnb$
$$CN = \Bigg|\frac{f'(x)}{f(x)}x\Bigg|$$
$$= \Bigg|\frac{2b^xlnb}{2b^x}x\Bigg|$$
$$CN=|x|lnb$$


8.Determine if the following series converges or diverges. If it converges determine its sum.
$$ \sum_{n=1}^\infty \frac{1}{2^n}$$
Listing the sequence gives us     
$$n=1,2,3,...$$
$$\frac{1}{2},\frac{1}{4},\frac{1}{8}, \frac{1}{16},...,\frac{1}{2^n}$$
We can see that the common ratio is $r=\frac{1}{2}$
Therefore:
$$S_n =\frac{a_1(1-r^n)}{1-r}$$
$$=\frac{\frac{1}{2}(1-\frac{1}{2}^n)}{1-\frac{1}{2}}$$
$$S_n=1-\frac{1^n}{2}$$
Now, the summation of the series 1 starting at 1 as n $\rightarrow \infty$ is equal to the limit of S as n $\rightarrow \infty$.
 $$ S_n = \frac{\frac{1}{2}(1)}{1-\frac{1}{2}} = \frac{\frac{1}{2}}{1-\frac{1}{2}} = \frac{\frac{1}{2}}{\frac{1}{2}} = 1$$
    $$lim_{n\rightarrow\infty} S_n = 1$$
   So, the series is convergent and its sum is 1.
